Improving the Reliability of the Pavlovian Go/No-Go Task

### 2023-04-19 manuscript/main.tex
checkpointing manuscript progress (2)

% p1: Summary of Pavlovian biases. [steal language from other papers]

Individual-differences in the expression of Pavlovian bias has been the focus of numerous published studies. Pavlovian biases have been observed to vary as a function of many psychiatric conditions. For example, a bias towards passive avoidance behavior has been observed for general and social anxiety \cite{mkrtchian2017modeling, peterburs2021impact}, but active avoidance is enhanced in individuals with a history of suicidal thoughts or behaviors \cite{millner2019suicidal}. Overall greater Pavlovian bias has been observed for trauma-exposure \cite{ousdal2018impact} and first-episode psychosis \cite{montagnese2020reinforcement}, but is attenuated in depression \cite{huys2016specificity} and schizophrenia \cite{albrecht2016reduction}. Pavlovian biases also vary across the lifespan as a function of age, with decreasing biases from childhood to young adulthood and recurs in older age \cite{raab2020adolescents, betts2020learning}. Pavlovian biases have been associated with individual differences in personality, such as impulsivity \cite{eisinger2020pavlovian}, and genetics \cite{richter2014valenced, richter2021motivational} (but not always; \cite{montagnese2020reinforcement}). Pavlovian biases also seem to be modulated by acute state effects, such as current mood \cite{weber2022effects}, anger \cite{wonderlich2020anger}, stress \cite{de2016acute}, and fear \cite{mkrtchian2017threat}.

To the best of our knowledge, none of the studies above investigated or reported the reliability of their behavioral measures. This is important because the reliability of a measure places an upper bound on the maximum observable correlation between itself and a second measure (e.g., symptom score). As an important corollary, as measure reliability decreases, so too does the power to detect individual-difference correlations \cite{Parsons2019-jw} thereby increasing the possibility of false-negative results (and simultaneously increasing the possibility of Type-S and -M errors; \cite{gelman2014beyond}). What studies do exist, however, unfortunately suggest that current behavioral measures from the Pavlovian go/no-go task are poor. Moutoussis and colleagues \cite{moutoussis2018change} investigated the psychometrics of the task in a large sample of adolescents and young adults. They found the test-retest reliability of model parameters to be poor. However, both their study population (adolescents) and retest intervals (6 - 18 mo) make it difficult to understand if the results reflect the study population, long retest intervals, or an interaction thereof. Recently, Pike and colleagues \cite{pike2022test} investigated the reliability of model parameters in an online adult sample. They too found poor estimates. Thus, the state of the current literature suggests that the task is poorly suited for investigating individual differences.

A second, no less important consideration is task repeatability, or that behavioral effects are robust to are repeated task administrations. One concern for task repeatability is practice effects, where participants’ performance on a task improves with repeated administrations. They might occur due to the attenuation of task-irrelevant nuisance factors (e.g., performance anxiety) and/or the learning of task-specific knowledge or strategies. Practice effects are relatively common for cognitive tasks [32, 33]. Practice effects are not inherently an issue for reliability — especially if an experimenter is only interested in the consistency, but not the absolute agreement, of participants’ performance over time — but they can become a pernicious issue if they are exhibited differentially across participants or if they are severe enough to induce ceiling effects. For example, Moutoussis and colleages report an overall reduction in the expression of Pavlovian biases in follow-up sessions. This may be a particular issue for studies investigating behavior change as a function of symptom change, either naturally or as part of treatment.

% p5: Summary of paper.

@@

The research reported in this manuscript was supported in part by the National Center for Advancing Translational Sciences (NCATS), a component of the National Institute of Health (NIH), under award number UL1TR003017 (ND, YN). The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health. SZ was supported by an NSF Graduate Research Fellowship. The funders had no role in study design, data collection and analysis, decision to publish or preparation of the manuscript.

@@

>>>>

Humans (and other animals) have an innate tendency to approach rewarding stimuli and shrink from punishing stimuli \cite{carver1994behavioral}. These inflexible Pavlovian biases can benefit or interfere with instrumental (i.e., action-outcome) learning depending on the context. For example, in a Pavlovian go/no-go task that orthogonalizes action (go, no-go) and valence (reward, punishment), human participants make more correct responses on Pavlovian-instrumental congruent trials where they must initiate action to receive reward or inhibit action to avoid punishment than on Pavlovian-instrumental incongruent trials where they must inhibit action to receive reward or initiate action to avoid punishment \cite{guitart2012go, guitart2014action}. The Pavlovian-instrumental interactions observed in task performance may arise during action selection (i.e., rewarding and punishing stimuli potentiate and inaction, respectively; \cite{guitart2012go}) and during learning (i.e., learning more rewards following action and punishments following inaction; \cite{swart2017catecholaminergic}).

Individual-differences in the expression of Pavlovian biases on the Pavlovian go/no-go task have been the focus of numerous studies. Pavlovian biases have been observed to vary as a function of many psychiatric conditions. For example, individuals with general and social anxiety show an increased bias towards passive avoidance behavior \cite{mkrtchian2017modeling, peterburs2021impact}, whereas active avoidance is amplified in individuals with a history of suicidal thoughts or behaviors \cite{millner2019suicidal}.  Pavlovian biases are enhanced for individuals with trauma exposure \cite{ousdal2018impact} and first-episode psychosis \cite{montagnese2020reinforcement}, but attenuated in individuals with depression \cite{huys2016specificity} and schizophrenia \cite{albrecht2016reduction}. Pavlovian biases have also been observed to vary parabolically across the lifespan, with decreasing bias from childhood to young adulthood which increases again in older age \cite{raab2020adolescents, betts2020learning}. Pavlovian biases have been associated with individual differences in personality \cite{eisinger2020pavlovian} and genetics \cite{richter2014valenced, richter2021motivational} (but see also \cite{montagnese2020reinforcement}). Pavlovian biases are also modulated by differences in acute state, such as current mood \cite{weber2022effects}, anger \cite{wonderlich2020anger}, stress \cite{de2016acute}, and fear \cite{mkrtchian2017threat}.

None of the individual-difference studies above, however, investigated or reported the reliability of their task measures. This is important because the reliability of a measure places an upper bound on the maximum observable correlation between itself and a second measure (e.g., symptom scores; \cite{zorowitz2023improving}). As an important corollary, as the reliability of a task measure decreases, so too does the power to detect individual-difference correlations, which increases the possibility of false-negative results \cite{Parsons2019-jw} and bias in effects that do reach statistical significance \cite{gelman2014beyond}. What studies that do exist suggest that the reliability of performance measures on the Pavlovian go/no-go task are poor. Moutoussis and colleagues \cite{moutoussis2018change} studied the psychometrics of the task in a large cohort of adolescents and young adults and found that both descriptive and model-based behavioral measures exhibited poor test-retest reliability. It was unclear, however, if this finding reflected the developmental sample and/or their lengthy test-retest intervals (6 and 18 months). Recently, Pike and colleagues \cite{pike2022test} investigated reliability of descriptive and model-based behavioral Pavlovian go/no-go task measures in an online adult sample over a 2-week test-retest interval. Regardless, they too found unacceptable reliability. Thus, the current state of the literature suggests that the Pavlovian go/no-go task is poorly suited for use in individual-differences studies.

A second psychometric consideration is task repeatability, or the ability of task to consistently elicit an expected pattern of behavior across repeated administrations. One concern for task repeatability is practice effects, where participants’ performance on a task improves with increasing exposure. Practice effects are relatively common for cognitive tasks \cite{hausknecht2007retesting, scharfen2018retest}. The repeatability of the Pavlovian go/no-go task has received less attention, though Moutoussis and colleagues reported that task performance improved and Pavlovian biases decreased over both 6 and 18 month follow-ups \cite{moutoussis2018change}. Practice effects can be confounding in longitudinal studies, especially for studies where an experimenter would expect reductions in bias (e.g., avoidance biases following a psychotherapy intervention). Moreover, practice effects can hamper diminish reliability if they manifest differentially across participants or if they minimize between-participant variability due to ceiling effects.

The objective of the present study was to investigate the reliability and repeatability of a gamified version of the Pavlovian go/no-go task. Gamification, or the incorporation of (video) game design elements into cognitive tasks, can increase participant engagement \cite{sailer2017gamification} and has previously been found to improve the reliability of task measures \cite{kucina2022solution, verdejo2021unified}. In Experiment 1, a sample of N=103 participants completed the gamified task four times over one month. We used reinforcement learning models quantify Pavlovian biases in choice behavior. Contrary to our expectations, we observed large practice effects in participants' performance, including the virtual disappearance of Pavlovian biases, that in turn impacted the reliability of estimated model parameters. In Experiment 2, a sample of N=110 participants completed a modified version of the Pavlovian go/no-go task --- one that prevents participants from using alternative strategies to complete the task --- three times over two weeks. Here we observed comparably smaller practice effects and, crucially, acceptable reliability of estimated model parameters.

@@

The authors are grateful to Daniel Bennett for helpful feedback on the task design. The research reported in this manuscript was supported in part by the National Center for Advancing Translational Sciences (NCATS), a component of the National Institute of Health (NIH), under award number UL1TR003017 (ND, YN). The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health. SZ was supported by an NSF Graduate Research Fellowship. The funders had no role in study design, data collection and analysis, decision to publish or preparation of the manuscript.

@@

\clearpage
\subsection*{Task appraisals}

\begin{table}[h!]
    \centering
    \begin{tabular}{lccccc}
    \toprule
               & Study &   Day 0 &      Day 3 &     Day 14 &     Day 28 \\
    \midrule
    Difficulty &  1 &  2.5 (1.1) &  2.2 (0.9) &  2.1 (1.0) &  1.8 (0.8) \\
               &  2 &  3.3 (1.1) &  3.0 (1.1) &  3.1 (1.1) &          - \\
    \midrule
    Fun        &  1 &  3.9 (0.9) &  4.0 (1.0) &  3.9 (0.9) &  4.1 (0.9) \\
               &  2 &  3.9 (0.9) &  4.1 (0.8) &  4.0 (0.8) &          - \\
    \midrule
    Clarity    &  1 &  4.8 (0.6) &  4.9 (0.4) &  5.0 (0.2) &  4.9 (0.3) \\
               &  2 &  4.9 (0.5) &  4.9 (0.2) &  4.9 (0.2) &          - \\
    \bottomrule
    \end{tabular}
    \caption{Mean (sd) of participants' ratings of task difficulty, fun, and clarity. All ratings were made on a 5-point Likert scale. Difficulty: ``How difficult was the task?'' (Very easy = 1, Very hard = 5); Fun: ``How fun was the task?'' (Very boring = 1, Very fun = 5); Clarity: ``How clear were the instructions?'' (Very confusing = 1, Very Clear = 5).}
    \label{tab:tabS02}
\end{table}
